---
title: "Darwin Core mapping"
subtitle: "For: Manual of the Alien Plants of Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This document describes how we map the checklist data to Darwin Core.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(digest)
```

# Read source data

The source data is maintained as an Excel file elsewhere. It is manually copied to this repository.

Read source data:
 
```{r}
raw_data <- read_excel(path = here("data", "raw", "checklist.xlsx")) 
```

Preview raw data:

```{r}
glimpse(raw_data)
```

# Preprocessing

## Add Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name (not kingdom).

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

input_data <-
  raw_data |>
  remove_empty("rows") |>
  clean_names() |>
  mutate(
    taxon_id = paste(
      "alien-plants-belgium",
      "taxon",
      vdigest(taxon, algo = "md5"),
      sep = ":"
    )
  )
```

## Process distribution information (adds rows)

1. Information on the occurrences is given for the **regions**, while date information is given for **Belgium** as a whole. Some transformations and clarifications are needed.

2. Some species have two values for `occurrenceStatus` and `eventDate`, i.e. species with the degree of naturalisation (`d_n`) of extinct (`Ext.`) or extinct/casual (`Ext./Cas.`).

 > Extinct: introduced taxa that once were naturalized (usually rather locally) but that have not been confirmed in recent times in their known localities. Only taxa that are certainly extinct are indicated as such.
 > Extinct/casual: Some of these extinct taxa are no longer considered as naturalized but still occur as casuals; such taxa are indicated as “Ext./Cas.” (for instance _Tragopogon porrifolius_).
 
For these species, we include the occurrenceStatus **within** the specified time frame (`eventDate` = first - most recent observation) and **after** the last observation (`eventDate` = most recent observation - current date).

The easiest way to do use a stepwise approach:

1. Clean presence information and date information in `distribution`
2. Create a separate dataframe `occurrenceStatus_ALO` (ALO = after last observation) for extinct and extinct/casual species
3. Map `occurrenceStatus` and `eventDate` from cleaned presence and date information in `distribution` (for `eventDate` = first - most recent observation)
4. Map `occurrenceStatus` and `eventDate` from cleaned presence and date information in `occurrence_status_ALO` (for `eventDate` = most recent observation - current date)
5. Bind both dataframes by row.
6. Map the other Darwin Core terms in the distribution extension

The checklist contains minimal presence information (`X`,`?` or `NA`) for the three regions in Belgium: Flanders, Wallonia and the Brussels-Capital Region, contained in `presence_fl`, `presence_wa` and `presence_br` respectively. Information regarding the first/last recorded observation applies to the distribution in Belgium as a whole. Both national and regional information is required in the checklist. In the `distribution.csv`, we will first provide `occurrenceStatus` and `eventDate`` on a **national level**, followed by specific information for the **regions**. 

For this, we use the following principles:

1. When a species is present in _only one region_, we can assume `eventDate` relates to that specific region. In this case, we can keep lines for Belgium and for the specific region populated with these variables (see #45).

2. When a species is present in _more than one_ region, it is impossible to extrapolate the date information for the regions. In this case, we decided to provide `occurrenceStatus` for the regional information, and specify dates only for Belgium.  

Thus, we need to specify when a species is present in only one of the regions.

We generate 4 new columns: `flanders`, `brussels`,`wallonia` and `belgium`. The content of these columns refers to the specific presence status of a species on a regional or national level. `S` if present in a single region or in Belgium, `?` if presence uncertain, `NA` if absent and `M` if present in multiple regions.

This should look like this:

```{r echo=FALSE}
as.data.frame(matrix(
  c(
    "X", NA, NA, "S", NA, NA, "S",
    NA, "X", NA, NA, "S", NA, "S", 
    NA, NA, "X", NA, NA, "S", "S",
    "X", "X", NA, "M", "M", NA, "S",
    "X", NA, "X", "M", NA, "M", "S",
    NA, "X", "X", NA, "M", "M", "S",
    NA, NA, NA, NA, NA, NA, NA,
    "X", "?", NA, "S", "?", NA, "S",
    "X", NA, "?", "S", NA, "?", "S",
    "X", "X", "?", "M", "M", "?", "S"
  ),
  ncol = 7,
  byrow = TRUE,
  dimnames = list(c(1:10), c(
    "presence_fl",
    "presence_br", 
    "presence_wa", 
    "flanders", 
    "brussels", 
    "wallonia",
    "belgium"
  ))
))
```

Add location columns:

```{r}
input_data <-
  input_data |>
  mutate(
    flanders = case_when(
      presence_fl == "X" &
      (is.na(presence_br) | presence_br == "?") &
      (is.na(presence_wa) | presence_wa == "?")
        ~ "S",
      presence_fl == "?" ~ "?",
      is.na(presence_fl) ~ NA_character_,
      TRUE ~ "M"
    ),
    brussels = case_when(
      (is.na(presence_fl) | presence_fl == "?") &
      presence_br == "X" &
      (is.na(presence_wa) | presence_wa == "?")
        ~ "S",
      presence_br == "?" ~ "?",
      is.na(presence_br) ~ NA_character_,
      TRUE ~ "M"
    ),
    wallonia = case_when(
      (is.na(presence_fl) | presence_fl == "?") &
      (is.na(presence_br) | presence_br == "?") &
      presence_wa == "X"
        ~ "S",
      presence_wa == "?" ~ "?",
      is.na(presence_wa) ~ NA_character_,
      TRUE ~ "M"
    ),
    belgium = case_when(
      presence_fl == "X" | presence_br == "X" | presence_wa == "X" ~ "S", # One is "X"
      presence_fl == "?" | presence_br == "?" | presence_wa == "?" ~ "?" # One is "?"
    )
  )
```

Summary of the previous action:

```{r}
input_data |>
  select(
    presence_fl,
    presence_br,
    presence_wa,
    flanders,
    wallonia,
    brussels,
    belgium) |>
  group_by_all() |>
  summarize(records = n()) |>
  arrange(flanders, wallonia, brussels)
```

Transform from wide to long (one line per distribution):

```{r}
input_data <-
  input_data |>
  pivot_longer(
    cols = c(flanders, wallonia, brussels, belgium),
    names_to = "location",
    values_to = "presence",
    values_drop_na = FALSE
  ) |>
  # Remove empty presence, but keep at least Belgium (for records without any distribution)
  filter(!is.na(presence) | location == "belgium")
```

Now we'll create and clean a start and end year:

```{r}
current_year = format(Sys.Date(), "%Y")

input_data <-
  input_data |>
  mutate(
    # start_year = first record (fr), end_year = most recent record (mrr)
    # Cleaning strips out ?, ca., < and >
    start_year = str_replace_all(fr, "(\\?|ca. |<|>)", ""),
    end_year = str_replace_all(mrr, "(\\?|ca. |<|>)", "")
  ) |>
  # If end_year = "Ann." or "N", use current year
  mutate(
    end_year = case_match(
      end_year,
      "Ann." ~ current_year,
      "N" ~ current_year,
      .default = end_year
    )
  )
```

Check if any `start_year` fall after `end_year` (expected to be none):

```{r}
input_data |>
  select(start_year, end_year) |>
  mutate(start_year = as.numeric(start_year)) |>
  mutate(end_year = as.numeric(end_year)) |>
  group_by(start_year, end_year) |>
  summarize(records = n()) |>
  filter(start_year > end_year) 
```

Combine `start_year` and `end_year` in an ranged `date` (ISO 8601 format). If any those two dates is empty or the same, we use a single year, as a statement when it was seen once (either as a first record or a most recent record):

```{r}
input_data <-
  input_data |>
  mutate(
    date = case_when(
      start_year == "" & end_year == "" ~ "",
      start_year == "" ~ end_year,
      end_year == "" ~ start_year,
      start_year == end_year ~ start_year,
      TRUE ~ paste(start_year, end_year, sep = "/")
    )
  )
```

Then, we map `occurrence_status` using [IUCN definitions](http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources) and set `event_date` only when `presence` = `S`:

```{r}
input_data <-
  input_data |>
  mutate(
    occurrence_status = case_match(
      presence,
      "S" ~ "present",
      "M" ~ "present",
      "?" ~ "presence uncertain",
      .default = ""
    ),
    event_date = case_when(
      presence == "S" ~ date,
      TRUE ~ ""
    )
  )
```

Overview of `occurrence_status` for each location x presence combination:

```{r}
input_data |>
  select (location, presence, occurrence_status) |>
  group_by_all() |>
  summarize(records = n()) 
```

Now we create `occurrence_status_ALO`:

```{r}
occurrence_status_ALO <-
  input_data |>
  filter(d_n == "Ext." | d_n == "Ext./Cas.")
```

Map `occurrence_status` and `event_date` for `occurrence_status_ALO`:

```{r}
occurrence_status_ALO <-
  occurrence_status_ALO |>
  mutate(
    occurrence_status = case_when(
      d_n == "Ext." ~ "absent",
      d_n == "Ext./Cas." ~ "present"
    ),
    event_date = case_when(
      presence == "S" ~ paste(end_year, current_year, sep = "/")
    )
  )
```

Bind `input_data` with `occurrence_status_ALO` by rows:

```{r}
input_data <- bind_rows(input_data, occurrence_status_ALO)
```

## Process pathways (adds rows)

Pathway information is based on `v_i`, which contains a list of introduction pathways (e.g. `Agric., wool`). We'll separate, clean, map and combine these values.

Split `v_i` on space in 4 columns. In case there are more than 4 values, these will be merged in native_range_4. The dataset currently contains no more than 4 values per record.

```{r}
input_data <-
  input_data |>
  separate(
    v_i,
    into = c("pathway_1", "pathway_2", "pathway_3", "pathway_4"),
    sep = ",",
    remove = FALSE,
    convert = FALSE,
    extra = "merge",
    fill = "right"
  )
```

Transform from wide to long (one line per distribution + pathway):

```{r}
input_data <-
  input_data |>
  pivot_longer(
    cols = c(pathway_1, pathway_2, pathway_3, pathway_4),
    names_to = "pathway_key",
    values_to = "pathway",
    values_drop_na = FALSE
  ) |>
  # Remove empty pathways, but keep at least first (for records without any pathway)
  filter(!is.na(pathway) | pathway_key == "pathway_1")
```

Map values:

```{r}
input_data <-
  input_data |>
  arrange(id) |> # To see descriptions in order
  mutate(
    pathway = str_replace_all(pathway, "\\?|…|\\.{3}", ""), # Strip ?, …, ...
    pathway = str_to_lower(pathway), # Convert to lowercase
    pathway = str_trim(pathway) # Clean whitespace
  ) |>
  mutate(
    type = "pathway",
    mapped_pathway = case_match(
      pathway,
      "agric." ~ "escape_agriculture",
      "bird seed" ~ "contaminant_seed",
      "birdseed" ~ "contaminant_seed",
      "bulbs" ~ "",
      "coconut mats" ~ "contaminant_seed",
      "fish" ~ "",
      "food refuse" ~ "escape_food_bait",
      "grain" ~ "contaminant_seed",
      "grain (rice)" ~ "contaminant_seed",
      "grass seed" ~ "contaminant_seed",
      "hay" ~ "",
      "hort" ~ "escape_horticulture",
      "hort." ~ "escape_horticulture",
      "hybridization" ~ "",
      "military troops" ~ "",
      "nurseries" ~ "contaminant_nursery",
      "ore" ~ "contaminant_habitat_material",
      "pines" ~ "contaminant_on_plants",
      "rice" ~ "",
      "salt" ~ "",
      "seeds" ~ "contaminant_seed",
      "timber" ~ "contaminant_timber",
      "tourists" ~ "stowaway_people_luggage",
      "traffic" ~ "",
      "unknown" ~ "unknown",
      "urban weed" ~ "stowaway",
      "waterfowl" ~ "contaminant_on_animals",
      "wool" ~ "contaminant_on_animals",
      "wool alien" ~ "contaminant_on_animals",
      .default = ""
    )
  )
```

Show mapped values:

```{r}
input_data |>
  select(pathway, mapped_pathway) |>
  group_by(pathway, mapped_pathway) |>
  summarize(records = n()) |>
  arrange(pathway)
```

## Process degree of establishment

The information for degree of establishment is contained in `d_n`. We decided to use the unified framework for biological invasions of [Blackburn et al. 2011](http://doc.rero.ch/record/24725/files/bach_puf.pdf) for `invasion stage`. `casual`, `naturalized` and `invasive` are terms included in this framework. However, we decided to discard the terms `naturalized` and `invasive` listed in Blackburn et al. (see trias-project/alien-fishes-checklist#6 (comment)). So, `naturalized` and `invasive` are replaced by `established`. For `extinct` (introduced taxa that once were naturalized but that have not been confirmed in recent times) and `extinct/casual` species (taxa are no longer considered as naturalized but still occur as casuals), we map the most recent invasion stage (i.e. `extinct` and `casual` respectively):

Map values:

```{r}
input_data <-
  input_data |>
  arrange(id) |> # To see descriptions in order
  mutate(
    degree_of_establishment = case_match(
      d_n,
      "Ext." ~ "extinct",
      "Ext.?" ~ "extinct",
      "Ext./Cas." ~ "casual",
      "Cas." ~ "casual",
      "Cas.?" ~ "casual",
      "Inv." ~ "established",
      "Nat." ~ "established",
      "Nat.?" ~ "established",
      .default = ""
    )
  )
```

Show mapped values:

```{r}
input_data |>
  select(d_n, degree_of_establishment) |>
  group_by(d_n, degree_of_establishment) |>
  summarize(records = n()) |>
  arrange(d_n)
```

# Darwin Core mapping

## Taxon core

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml):

```{r}
taxon <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |> # Only keep one row per taxon
  mutate(
    .keep = "none",
    language = "en",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    rightsHolder = "Meise Botanic Garden",
    datasetID = "https://doi.org/10.15468/wtda1m",
    datasetName = "Manual of the Alien Plants of Belgium",
    taxonID = taxon_id,
    scientificNameID = scientific_name_id,
    scientificName = taxon,
    kingdom = "Plantae",
    family = family,
    taxonRank = taxon_rank,
    nomenclaturalCode = "ICN"
  ) |>
  relocate(family, .after = kingdom) |>
  arrange(taxonID)
```

## Distribution extension

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml):

```{r}
distribution <-
  input_data |>
  distinct(taxon_id, location, event_date, .keep_all = TRUE) |> # Only keep one row per distribution
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    locationID = case_match(
      location,
      "belgium" ~ "ISO_3166-2:BE",
      "flanders" ~ "ISO_3166-2:BE-VLG",
      "wallonia" ~ "ISO_3166-2:BE-WAL",
      "brussels" ~ "ISO_3166-2:BE-BRU"
    ),
    locality = case_match(
      location,
      "belgium" ~ "Belgium",
      "flanders" ~ "Flemish Region",
      "wallonia" ~ "Walloon Region",
      "brussels" ~ "Brussels-Capital Region"
    ),
    countryCode = "BE",
    occurrenceStatus = occurrence_status,
    establishmentMeans = "introduced",
    eventDate = event_date
  ) |>
  arrange(taxonID)
```

## Species profile extension

In this extension will express broad habitat characteristics (e.g. `isTerrestrial`) of the species. Habitat information can be found in `habitat`, which describes whether a species is found in freshwater, terrestrial or both (terrestrial/freshwater) habitats.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml):

```{r}
species_profile <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |> # Only keep one row per taxon
  mutate(habitat = str_to_lower(habitat)) |>
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    isMarine = "FALSE",
    isFreshwater = case_when(
      habitat == "fresh water" | habitat == "terrestrial/fresh water" ~ "TRUE",
      TRUE ~ "FALSE"
    ),
    isTerrestrial = case_when(
      habitat == "terrestrial" | habitat == "terrestrial/fresh water" ~ "TRUE",
      TRUE ~ "FALSE"
    )
  ) |>
  arrange(taxonID)
```

## Description extension

In the description extension we include the native range information. This information is included in `origin` (e.g. `E AS-Te NAM`). We'll separate, clean, map and combine these values.

Split `origin` on space in 4 columns. In case there are more than 4 values, these will be merged in native_range_4. The dataset currently contains no more than 4 values per record.

```{r}
native_range <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |> # Only keep one row per taxon
  separate(
    origin,
    into = c("native_range_1", "native_range_2", "native_range_3", "native_range_4"),
    sep = " ",
    remove = FALSE,
    convert = FALSE,
    extra = "merge",
    fill = "right"
  )
```

Transform from wide to long (one line per native range):

```{r}
native_range <-
  native_range |>
  pivot_longer(
    cols = c(native_range_1, native_range_2, native_range_3, native_range_4),
    names_to = "native_range_key",
    values_to = "native_range",
    values_drop_na = TRUE # Also removes records for which there is no native_range_1
  )
```

Map values:

```{r}
native_range <-
  native_range |>
  arrange(id) |> # To see descriptions in order
  mutate(
    native_range = str_replace_all(native_range, "\\?", ""), # Strip ?
    native_range = str_trim(native_range) # Clean whitespace
  ) |>
  mutate(
    mapped_native_range = case_match(
      native_range,
      "AF" ~ "Africa (WGSRPD:2)",
      "AM" ~ "pan-American",
      "AS" ~ "Asia",
      "AS-Te" ~ "temperate Asia (WGSRPD:3)",
      "AS-Tr" ~ "tropical Asia (WGSRPD:4)",
      "AUS" ~ "Australasia (WGSRPD:5)",
      "Cult." ~ "cultivated origin",
      "E" ~ "Europe (WGSRPD:1)",
      "Hybr." ~ "hybrid origin",
      "NAM" ~ "Northern America (WGSRPD:7)",
      "SAM" ~ "Southern America (WGSRPD:8)",
      "Trop." ~ "Pantropical",
      .default = ""
    )
  ) |>
  filter(!is.na(mapped_native_range) & mapped_native_range != "")
```

Show mapped values:

```{r}
native_range |>
  select(native_range, mapped_native_range) |>
  group_by(native_range, mapped_native_range) |>
  summarize(records = n()) |>
  arrange(native_range)
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml):

```{r}
description <-
  native_range |>
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    description = mapped_native_range,
    type = "native range",
    language = "en"
  ) |>
  relocate(type, .after = description) |>
  arrange(taxonID)
```

# Save to CSV

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
