---
title: "Darwin Core mapping"
subtitle: "For: Manual of the Alien Plants of Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This document describes how we map the checklist data to Darwin Core.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(magrittr)       # To use %<>% pipes
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(digest)
```

# Read source data

The source data is maintained as an Excel file elsewhere. It is manually copied to this repository.

Read source data:
 
```{r}
raw_data <- read_excel(path = here("data", "raw", "checklist.xlsx")) 
```

Preview raw data:

```{r}
glimpse(raw_data)
```

# Preprocessing

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name (not kingdom).

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

input_data <-
  raw_data |>
  remove_empty("rows") |>
  clean_names() |>
  mutate(
    taxon_id = paste(
      "alien-plants-belgium",
      "taxon",
      vdigest(taxon, algo = "md5"),
      sep = ":"
    )
  )
```

# Darwin Core mapping

## Taxon core

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml):

```{r}
taxon <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |>
  mutate(
    .keep = "none",
    language = "en",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    rightsHolder = "Meise Botanic Garden",
    datasetID = "https://doi.org/10.15468/wtda1m",
    datasetName = "Manual of the Alien Plants of Belgium",
    taxonID = taxon_id,
    scientificNameID = scientific_name_id,
    scientificName = taxon,
    kingdom = "Plantae",
    family = family,
    taxonRank = taxon_rank,
    nomenclaturalCode = "ICN"
  ) |>
  relocate(family, .after = kingdom) |>
  arrange(taxonID)
```

## Distribution extension

Before we start mapping the distribution extension, we focus on two terms: `occurrenceStatus` and  `eventDate`. That is because:

1. Information on the occurrences is given for the **regions**, while date information is given for **Belgium** as a whole. Some transformations and clarifications are needed.

2. Some species have two values for `occurrenceStatus` and `eventDate`, i.e. species with the degree of naturalisation (`d_n`) of extinct (`Ext.`) or extinct/casual (`Ext./Cas.`).

 > Extinct: introduced taxa that once were naturalized (usually rather locally) but that have not been confirmed in recent times in their known localities. Only taxa that are certainly extinct are indicated as such.
 > Extinct/casual: Some of these extinct taxa are no longer considered as naturalized but still occur as casuals; such taxa are indicated as “Ext./Cas.” (for instance _Tragopogon porrifolius_).
 
For these species, we include the occurrenceStatus **within** the specified time frame (`eventDate` = first - most recent observation) and **after** the last observation (`eventDate` = most recent observation - current date).

The easiest way to do use a stepwize approach:

1. Clean presence information and date information in `distribution`
2. Create a separate dataframe `occurrenceStatus_ALO` (ALO = after last observation) for extinct and extinct/casual species
3. Map `occurrenceStatus` and `eventDate` from cleaned presence and date information in `distribution` (for `eventDate` = first - most recent observation)
4. Map `occurrenceStatus` and `eventDate` from cleaned presence and date information in `occurrence_status_ALO` (for `eventDate` = most recent observation - current date)
5. Bind both dataframes by row.
6. Map the other Darwin Core terms in the distribution extension

The checklist contains minimal presence information (`X`,`?` or `NA`) for the three regions in Belgium: Flanders, Wallonia and the Brussels-Capital Region, contained in `presence_fl`, `presence_wa` and `presence_br` respectively. Information regarding the first/last recorded observation applies to the distribution in Belgium as a whole. Both national and regional information is required in the checklist. In the `distribution.csv`, we will first provide `occurrenceStatus` and `eventDate`` on a **national level**, followed by specific information for the **regions**. 

For this, we use the following principles:

1. When a species is present in _only one region_, we can assume `eventDate` relates to that specific region. In this case, we can keep lines for Belgium and for the specific region populated with these variables (see #45).

2. When a species is present in _more than one_ region, it is impossible to extrapolate the date information for the regions. In this case, we decided to provide `occurrenceStatus` for the regional information, and specify dates only for Belgium.  

Thus, we need to specify when a species is present in only one of the regions.

We generate 4 new columns: `flanders`, `brussels`,`wallonia` and `belgium`. The content of these columns refers to the specific presence status of a species on a regional or national level. `S` if present in a single region or in Belgium, `?` if presence uncertain, `NA` if absent and `M` if present in multiple regions.

This should look like this:

```{r echo=FALSE}
as.data.frame(matrix(
  c(
    "X", NA, NA, "S", NA, NA, "S",
    NA, "X", NA, NA, "S", NA, "S", 
    NA, NA, "X", NA, NA, "S", "S",
    "X", "X", NA, "M", "M", NA, "S",
    "X", NA, "X", "M", NA, "M", "S",
    NA, "X", "X", NA, "M", "M", "S",
    NA, NA, NA, NA, NA, NA, NA,
    "X", "?", NA, "S", "?", NA, "S",
    "X", NA, "?", "S", NA, "?", "S",
    "X", "X", "?", "M", "M", "?", "S"
  ),
  ncol = 7,
  byrow = TRUE,
  dimnames = list(c(1:10), c(
    "presence_fl",
    "presence_br", 
    "presence_wa", 
    "flanders", 
    "brussels", 
    "wallonia",
    "belgium"
  ))
))
```

We translate this to the distribution extension:

```{r}
input_distribution <-
  input_data |>
  mutate(
    flanders = case_when(
      presence_fl == "X" &
      (is.na(presence_br) | presence_br == "?") &
      (is.na(presence_wa) | presence_wa == "?")
        ~ "S",
      presence_fl == "?" ~ "?",
      is.na(presence_fl) ~ "NA",
      TRUE ~ "M"
    ),
    brussels = case_when(
      (is.na(presence_fl) | presence_fl == "?") &
      presence_br == "X" &
      (is.na(presence_wa) | presence_wa == "?")
        ~ "S",
      presence_br == "?" ~ "?",
      is.na(presence_br) ~ "NA",
      TRUE ~ "M"
    ),
    wallonia = case_when(
      (is.na(presence_fl) | presence_fl == "?") &
      (is.na(presence_br) | presence_br == "?") &
      presence_wa == "X"
        ~ "S",
      presence_wa == "?" ~ "?",
      is.na(presence_wa) ~ "NA",
      TRUE ~ "M"
    ),
    belgium = case_when(
      presence_fl == "X" | presence_br == "X" | presence_wa == "X" ~ "S", # One is "X"
      presence_fl == "?" | presence_br == "?" | presence_wa == "?" ~ "?" # One is "?"
    )
  )
```

Summary of the previous action:

```{r}
input_distribution |>
  select(
    presence_fl,
    presence_br,
    presence_wa,
    flanders,
    wallonia,
    brussels,
    belgium) |>
  group_by_all() |>
  summarize(records = n()) |>
  arrange(flanders, wallonia, brussels)
```

One line should represent the presence information of a species in one region or Belgium. We need to transform `input_distribution` from a wide to a long table (i.e. create a `location` and `presence` column):

```{r}
input_distribution <-
  input_distribution |>
  gather(
    key, value,
    flanders, wallonia, brussels, belgium,
    convert = FALSE
  ) |>
  rename("location" = "key", "presence" = "value") |>
  filter(!presence == "NA") # Remove species which lack presence information
```

Now we'll create and clean a start and end year:

```{r}
current_year = format(Sys.Date(), "%Y")

input_distribution <-
  input_distribution |>
  mutate(
    # start_year = first record (fr), end_year = most recent record (mrr)
    # Cleaning strips out ?, ca., < and >
    start_year = str_replace_all(fr, "(\\?|ca. |<|>)", ""),
    end_year = str_replace_all(mrr, "(\\?|ca. |<|>)", "")
  ) |>
  # If end_year = "Ann." or "N", use current year
  mutate(
    end_year = case_match(
      end_year,
      "Ann." ~ current_year,
      "N" ~ current_year,
      .default = end_year
    )
  )
```

Check if any `start_year` fall after `end_year` (expected to be none):

```{r}
input_distribution |>
  select(start_year, end_year) |>
  mutate(start_year = as.numeric(start_year)) |>
  mutate(end_year = as.numeric(end_year)) |>
  group_by(start_year, end_year) |>
  summarize(records = n()) |>
  filter(start_year > end_year) 
```

Combine `start_year` and `end_year` in an ranged `date` (ISO 8601 format). If any those two dates is empty or the same, we use a single year, as a statement when it was seen once (either as a first record or a most recent record):

```{r}
input_distribution <-
  input_distribution |>
  mutate(
    date = case_when(
      start_year == "" & end_year == "" ~ "",
      start_year == "" ~ end_year,
      end_year == "" ~ start_year,
      start_year == end_year ~ start_year,
      TRUE ~ paste(start_year, end_year, sep = "/")
    )
  )
```

Then, we map `occurrence_status` using [IUCN definitions](http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources) and set `event_date` only when `presence` = `S`:

```{r}
input_distribution <-
  input_distribution |>
  mutate(
    occurrence_status = case_match(
      presence,
      "S" ~ "present",
      "M" ~ "present",
      "?" ~ "presence uncertain",
      .default = ""
    ),
    event_date = case_when(
      presence == "S" ~ date,
      TRUE ~ ""
    )
  )
```

Overview of `occurrence_status` for each location x presence combination:

```{r}
input_distribution |>
  select (location, presence, occurrence_status) |>
  group_by_all() |>
  summarize(records = n()) 
```

Now we create `occurrence_status_ALO`:

```{r}
occurrence_status_ALO <-
  input_distribution |>
  filter(d_n == "Ext." | d_n == "Ext./Cas.")
```

Map `occurrence_status` and `event_date` for `occurrence_status_ALO`:

```{r}
occurrence_status_ALO <-
  occurrence_status_ALO |>
  mutate(
    occurrence_status = case_when(
      d_n == "Ext." ~ "absent",
      d_n == "Ext./Cas." ~ "present"
    ),
    event_date = case_when(
      presence == "S" ~ paste(end_year, current_year, sep = "/")
    )
  )
```

Bind `occurrence_status_ALO` and `input_distribution` by rows:

```{r}
input_distribution <- bind_rows(input_distribution, occurrence_status_ALO)
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml):

```{r}
distribution <-
  input_distribution |>
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    locationID = case_match(
      location,
      "belgium" ~ "ISO_3166-2:BE",
      "flanders" ~ "ISO_3166-2:BE-VLG",
      "wallonia" ~ "ISO_3166-2:BE-WAL",
      "brussels" ~ "ISO_3166-2:BE-BRU"
    ),
    locality = case_match(
      location,
      "belgium" ~ "Belgium",
      "flanders" ~ "Flemish Region",
      "wallonia" ~ "Walloon Region",
      "brussels" ~ "Brussels-Capital Region"
    ),
    countryCode = "BE",
    occurrenceStatus = occurrence_status,
    establishmentMeans = "introduced",
    eventDate = event_date
  ) |>
  arrange(taxonID)
```

## Species profile extension

In this extension will express broad habitat characteristics (e.g. `isTerrestrial`) of the species.

```{r}
species_profile <- input_data
```

Habitat information can be found in `habitat`, which describes whether a species is found in freshwater, terrestrial or both (terrestrial/freshwater) habitats.

Show unique values:

```{r}
species_profile %>%
  distinct(habitat) %>%
  arrange(habitat)
```

Clean content somewhat, i.e. change uppercase to lowercase

```{r}
species_profile %<>% mutate(habitat = str_to_lower(habitat)) 
```

We map this information respectively to `isFreshwater`, `isTerrestrial` or both terms in the species profile extension.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml):

### taxonID

```{r}
species_profile %<>% mutate(dwc_taxonID = taxon_id)
```

### isMarine

```{r}
species_profile %<>% mutate(dwc_isMarine = "FALSE")
```

### isFreshwater

```{r}
species_profile %<>% mutate(dwc_isFreshwater = case_when(
  habitat == "fresh water" | habitat == "terrestrial/fresh water" ~ "TRUE",
  TRUE ~"FALSE"
))
```

### isTerrestrial

```{r}
species_profile %<>% mutate(dwc_isTerrestrial = case_when(
  habitat == "terrestrial" | habitat == "terrestrial/fresh water" ~ "TRUE",
  TRUE ~"FALSE"
))
```

Show mapped values:

```{r}
species_profile %>%
  select(habitat, dwc_isMarine, dwc_isFreshwater, dwc_isTerrestrial) %>%
  group_by_all() %>%
  summarize(records = n())
```

## Description extension

In the description extension we want to include several important characteristics (hereafter refered to as descriptors) about the species:

- Native range
- Pathway of introduction
- Invasion stage

A single taxon can have multiple descriptions of the same type (e.g. multiple native ranges), expressed as multiple rows in the description extension.

For each descriptor, we create a separate dataframe to process the specific information. We always specify _which descriptor_ we map (`type` column) and its _specific content_ (`description` column). After the mapping of these Darwin Core terms `type` and `value`, we merge the dataframes to generate one single description extension. We then continue the mapping process by adding the other Darwin Core terms (which content is independent of the type of descriptor, such as `language`).

### native range
 
`origin` contains native range information (e.g. `E AS-Te NAM`). We'll separate, clean, map and combine these values.

Create separate dataframe:

```{r}
native_range <- input_data
```

Create `description` from `origin`:

```{r}
native_range %<>% mutate(description = origin)
```

Separate `description` on space in 4 columns:
In case there are more than 4 values, these will be merged in native_range_4. 
The dataset currently contains no more than 4 values per record.

```{r}
native_range %<>% separate(
  description,
  into = c("native_range_1", "native_range_2", "native_range_3", "native_range_4"),
  sep = " ",
  remove = TRUE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather native ranges in a key and value column:

```{r}
native_range %<>% gather(
  key, value,
  native_range_1, native_range_2, native_range_3, native_range_4,
  na.rm = TRUE, # Also removes records for which there is no native_range_1
  convert = FALSE
)
```

Sort on ID to see pathways in context for each record:

```{r}
native_range %<>% arrange(id)
```

Clean values:

```{r}
native_range %<>% mutate(
  value = str_replace_all(value, "\\?", ""), # Strip ?
  value = str_trim(value) # Clean whitespace
)
```

Map values:

```{r}
native_range %<>% mutate(mapped_value = recode(value,
  "AF" = "Africa (WGSRPD:2)",
  "AM" = "pan-American",
  "AS" = "Asia",
  "AS-Te" = "temperate Asia (WGSRPD:3)",
  "AS-Tr" = "tropical Asia (WGSRPD:4)",
  "AUS" = "Australasia (WGSRPD:5)",
  "Cult." = "cultivated origin",
  "E" = "Europe (WGSRPD:1)",
  "Hybr." = "hybrid origin",
  "NAM" = "Northern America (WGSRPD:7)",
  "SAM" = "Southern America (WGSRPD:8)",
  "Trop." = "Pantropical",
  .default = "",
  .missing = "" # As result of stripping, records with no native range already removed by gather()
))
```

Show mapped values:

```{r}
native_range %>%
  select(value, mapped_value) %>%
  group_by(value, mapped_value) %>%
  summarize(records = n()) %>%
  arrange(value)
```

Drop `key` and `value` column and rename `mapped value`:

```{r}
native_range %<>% select(-key, -value)
native_range %<>% rename(description = mapped_value)
```

Keep only non-empty descriptions:

```{r}
native_range %<>% filter(!is.na(description) & description != "")
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

### pathway of introduction

Pathway information is based on `v_i`, which contains a list of introduction pathways (e.g. `Agric., wool`). We'll separate, clean, map and combine these values.

Create separate dataframe:

```{r}
pathway <- input_data
```

Create `pathway` from `v_i`:

```{r}
pathway %<>% mutate(pathway = v_i)
```

Separate `pathway` on `,` in 4 columns:
In case there are more than 4 values, these will be merged in pathway_4. 
The dataset currently contains no more than 3 values per record.

```{r}
pathway %<>% separate(
  pathway,
  into = c("pathway_1", "pathway_2", "pathway_3", "pathway_4"),
  sep = ",",
  remove = TRUE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather pathways in a key and value column:

```{r}
pathway %<>% gather(
  key, value,
  pathway_1, pathway_2, pathway_3, pathway_4,
  na.rm = TRUE, # Also removes records for which there is no pathway_1
  convert = FALSE
)
```

Sort on `taxon_id` to see pathways in context for each record:

```{r}
pathway %<>% arrange(taxon_id)
```

Show unique values:

```{r}
pathway %>%
  distinct(value) %>%
  arrange(value) 
```

Clean values:

```{r}
pathway %<>% mutate(
  value = str_replace_all(value, "\\?|…|\\.{3}", ""), # Strip ?, …, ...
  value = str_to_lower(value), # Convert to lowercase
  value = str_trim(value) # Clean whitespace
)
```

Map values to the CBD standard::

```{r}
pathway %<>% mutate(cbd_stand = recode(value, 
  "agric." = "escape_agriculture",
  "bird seed" = "contaminant_seed",
  "birdseed" = "contaminant_seed",
  "bulbs" = "",
  "coconut mats" = "contaminant_seed",
  "fish" = "",
  "food refuse" = "escape_food_bait",
  "grain" = "contaminant_seed",
  "grain (rice)" = "contaminant_seed",
  "grass seed" = "contaminant_seed",
  "hay" = "",
  "hort" = "escape_horticulture",
  "hort." = "escape_horticulture",
  "hybridization" = "",
  "military troops" = "",
  "nurseries" = "contaminant_nursery",
  "ore" = "contaminant_habitat_material",
  "pines" = "contaminant_on_plants",
  "rice" = "",
  "salt" = "",
  "seeds" = "contaminant_seed",
  "timber" = "contaminant_timber",
  "tourists" = "stowaway_people_luggage",
  "traffic" = "",
  "unknown" = "unknown",
  "urban weed" = "stowaway",
  "waterfowl" = "contaminant_on_animals",
  "wool" = "contaminant_on_animals",
  "wool alien" = "contaminant_on_animals",
  .default = "",
  .missing = "" # As result of stripping, records with no pathway already removed by gather()
))
```

Add prefix `cbd_2014_pathway` in case there is a match with the CBD standard:

```{r}
pathway %<>% mutate(mapped_value = case_when(
  cbd_stand != "" ~ paste ("cbd_2014_pathway", cbd_stand, sep = ":"),
  TRUE ~ ""
))
```

Show mapped values:

```{r}
pathway %>%
  select(value, mapped_value) %>%
  group_by(value, mapped_value) %>%
  summarize(records = n()) %>%
  arrange(value) 
```

Drop `key`,`value` and `cbd_stand` column:

```{r}
pathway %<>% select(-key, -value, -cbd_stand)
```

Change column name `mapped_value` to `description`:

```{r}
pathway %<>%  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
pathway %<>% mutate (type = "pathway")
```

Show pathway descriptions:

```{r}
pathway %>% 
  select(description) %>% 
  group_by(description) %>% 
  summarize(records = n())
```

Keep only non-empty descriptions:

```{r}
pathway %<>% filter(!is.na(description) & description != "")
```

### Degree of establishment

Create separate dataframe:

```{r start_invasion_stage}
degree_of_establishment <- input_data
```

The information for degree of establishment is contained in `d_n`:

```{r}
degree_of_establishment %>%
  select(d_n) %>%
  group_by_all() %>%
  summarize(records = n()) 
```

Clean the data:

```{r}
degree_of_establishment %<>% mutate(description = recode(d_n,
  "Ext.?" = "Ext.",
  "Cas.?" = "Cas.",
  "Nat.?" = "Nat.",
  .missing = ""
))
```

We decided to use the unified framework for biological invasions of [Blackburn et al. 2011](http://doc.rero.ch/record/24725/files/bach_puf.pdf) for `invasion stage`. `casual`, `naturalized` and `invasive` are terms included in this framework. However, we decided to discard the terms `naturalized` and `invasive` listed in Blackburn et al. (see trias-project/alien-fishes-checklist#6 (comment)). So, `naturalized` and `invasive` are replaced by `established`. For `extinct` (introduced taxa that once were naturalized but that have not been confirmed in recent times) and `extinct/casual` species (taxa are no longer considered as naturalized but still occur as casuals), we map the most recent invasion stage (i.e. `extinct` and `casual` respectively):

```{r}
degree_of_establishment %<>% mutate(description = recode(description,
  "Cas." = "casual",
  "Inv." = "established",
  "Nat." = "established",
  "Ext." = "extinct",
  "Ext./Cas." = "casual"
))
```

Show mapped values:

```{r}
degree_of_establishment %>%
  select(d_n, description) %>%
  group_by_all() %>%
  summarize(records = n())
```

Create a `type` field to indicate the type of description:

```{r}
degree_of_establishment %<>% mutate(type = "degree of establishment")
```

Union native range, pathway of introduction and degree of establishment:

```{r}
description <- bind_rows(native_range, pathway, degree_of_establishment)
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml):

### taxonID

```{r}
description %<>% mutate(dwc_taxonID = taxon_id)
```

### description

```{r}
description %<>% mutate(dwc_description = description)
```

### type

```{r}
description %<>% mutate(dwc_type = type)
```

### language

```{r}
description %<>% mutate(dwc_language = "en")
```

# Post-processing

Remove the original columns for each of the generated files:

```{r}
distribution %<>% select(starts_with("dwc_")) 
species_profile %<>% select(starts_with("dwc_")) 
description %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description) <- str_remove(colnames(description), "dwc_")
```

Sort on `taxonID` to group description information per taxon:

```{r}
species_profile %<>% arrange(taxonID)
description %<>% arrange(taxonID)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
